---
title: "Fire Regime Summary"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Fire Regime types 

This report outlines a summary of fire regime types across the province and across the BUMO Study area. For this analysis we used all historic fire perimeters from the bcdata catalogue (1917 - 2025). The fire regime was built based on methodology documented by Morgan et al. Analysis details are found [here](https://github.com/BuMo-Wildfire-Resilience/BuMo_data_prep/blob/main/UBC_FireRegime.R)


```{r input information , echo = FALSE, warning = FALSE, message = FALSE}
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
library(openxlsx)
library(knitr)

DataDir <- 'data'
spatialDir <- fs::path(DataDir,'spatial')

OutDir <- 'out'
dataOutDir <- file.path(OutDir,'data')
spatialOutDir <- file.path(OutDir,'spatial')

#aoi <-  st_read(file.path(spatialDir, "AOI_50k.gpkg"))
aoi <- st_read(file.path(spatialOutDir, "AOI_Admin.gpkg"), quiet = TRUE)

# read in the key 
key <- read.xlsx(fs::path(spatialDir, "fire_regime", "BUMO_regimeT.xlsx"))
key <- key |> 
  mutate(MAP_LABEL = BEC_variant)

```

We can review the proportion of the province in which the fire regimes are located. 

The table below shows the area of each fire regime type across the province and the proportion of each fire regime as a percentage of total area. The NA value represents those BEC units which are not assigned to a BUMO_regime, likley as they do not occur within the BUMO study area. 

```{r, echo = FALSE}

#bec_sum <- read.csv( fs::path(spatialDir, "fire_regime","bec_fr_total_ha_BC.csv"))
fr_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_BC.csv")) |> 
  select(-X) |> 
  #rename("fr_pc_bc" = fr_pc_bumo) |> 
  select(BUMO_regime, fr_area_ha, total_ha, fr_pc_bc)

knitr::kable(fr_sum)

#fr_bumo_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_bumo.csv"))

```

The table below shows the area and percentage of each fire regime type within the BUMO study area. 

```{r, echo = FALSE}

fr_bumo_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_bumo.csv"))|> 
  select(-X) 

knitr::kable(fr_bumo_sum)

```


## 2. Fires across BC 

This plot shows the total area burnt across BC for all fire years, the number above each point indicates the total number of fires within that year. Note some early records are not accurate for fire numbers. 

```{r pressure, echo=FALSE, fig.cap="", out.width = '90%'}
p1 <- fs::path(spatialDir, "fire_regime","1_fires_across_bc.jpg")
knitr::include_graphics(p1)

```


We can review this based on the total area burnt (figure 2), or by the area in which a fire regime burnt relative to its total area across the province (figure 3)

Across the province we can determine what proportion of area burnt was made up of each fire regime types.

```{r plot2, echo=FALSE, fig.cap="Figure 2", out.width = '90%'}
p2 <- fs::path(spatialDir, "fire_regime","2_fire_fr_pc_across_bc.jpg")
knitr::include_graphics(p2)

```



```{r plot2a, echo=FALSE, fig.cap="Figure 3", out.width = '90%'}
p2a <- fs::path(spatialDir, "fire_regime","2a_fire_fr_pc__of_total_fr_across_bc.jpg")
knitr::include_graphics(p2a)

```

We can also review the proportions of the fire regimes burn within the BUMO study area 


```{r plot 4, echo=FALSE, fig.cap="Figure 4", out.width = '90%'}
p3 <- fs::path(spatialDir, "fire_regime","3_fire_fr_pc_across_bumo.jpg")
knitr::include_graphics(p3)
```














## Appendices 

The key to distinguish fire regimes based on BEC zones are shown in the table below. 

```{r fire_regime key, echo=FALSE}
knitr::kable(key)

```


