---
title: "Fire Regime Summary"
output: html_document
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Fire Regime types 

This report outlines a summary of fire regime types across the province and across the BUMO Study area. For this analysis we used all historic fire perimeters from the bcdata catalogue (1917 - 2025). The fire regime was built based on methodology documented by Morgan et al. Analysis details are found [here](https://github.com/BuMo-Wildfire-Resilience/BuMo_data_prep/blob/main/UBC_FireRegime.R)



Using the historic fire perimeters, we can plot the number of fires within the study area (aoi) + plus the buffered study area (using a expanded perimeter of 50km). 


Using the historic fire perimeters, we can plot the number of fires within the study area (aoi) + plus the buffered study area (using a expanded perimeter of 50km). 


```{r input information , echo = FALSE, warning = FALSE, message = FALSE}
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
library(openxlsx)
library(knitr)

DataDir <- 'data'
spatialDir <- fs::path(DataDir,'spatial')

OutDir <- 'out'
dataOutDir <- file.path(OutDir,'data')
spatialOutDir <- file.path(OutDir,'spatial')

#aoi <-  st_read(file.path(spatialDir, "AOI_50k.gpkg"))
aoi <- st_read(file.path(spatialOutDir, "AOI_Admin.gpkg"), quiet = TRUE)

# read in the key 
key <- read.xlsx(fs::path(spatialDir, "fire_regime", "BUMO_regimeT.xlsx"))
key <- key |> 
  mutate(MAP_LABEL = BEC_variant)

```

We can review the proportion of the province in which the fire regimes are located. 

The table below shows the area of each fire regime type across the province and the proportion of each fire regime as a percentage of total area. The NA value represents those BEC units which are not assigned to a BUMO_regime, likley as they do not occur within the BUMO study area. 

```{r, echo = FALSE}

#bec_sum <- read.csv( fs::path(spatialDir, "fire_regime","bec_fr_total_ha_BC.csv"))
fr_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_BC.csv")) |> 
  select(-X) |> 
  rename("fr_pc_bc" = fr_pc_bumo) |> 
  select(BUMO_regime, fr_area_ha, total_ha, fr_pc_bc)

knitr::kable(fr_sum)

#fr_bumo_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_bumo.csv"))

```

The table below shows the area within the BUMO study area with the same metrics

```{r, echo = FALSE}

fr_bumo_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_bumo.csv"))|> 
  select(-X) 

knitr::kable(fr_bumo_sum)

```















## Appendices 

The key to distinguish fire regimes based on BEC zones are shown in the table below. 

```{r fire_regime key, echo=FALSE}
knitr::kable(key)

```


