---
title: "Fire Regime Summary"
output:
  word_document: default
  html_document: default
date: "2026-02-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 1. Fire Regime types 

This report outlines a summary of fire regime types across the province and across the BUMO Study area. For this analysis we used all historic fire perimeters from the bcdata catalogue (1917 - 2025). The fire regime was built based on methodology documented by Morgan et al. Analysis details are found [here](https://github.com/BuMo-Wildfire-Resilience/BuMo_data_prep/blob/main/UBC_FireRegime.R)


```{r input information , echo = FALSE, warning = FALSE, message = FALSE}
library(sf)
library(terra)
library(dplyr)
library(ggplot2)
library(openxlsx)
library(knitr)

DataDir <- 'data'
spatialDir <- fs::path(DataDir,'spatial')

OutDir <- 'out'
dataOutDir <- file.path(OutDir,'data')
spatialOutDir <- file.path(OutDir,'spatial')

#aoi <-  st_read(file.path(spatialDir, "AOI_50k.gpkg"))
aoi <- st_read(file.path(spatialOutDir, "AOI_Admin.gpkg"), quiet = TRUE)

# read in the key 
key <- read.xlsx(fs::path(spatialDir, "fire_regime", "BUMO_regimeT.xlsx"))
key <- key |> 
  mutate(MAP_LABEL = BEC_variant)

```

We can review the proportion of the province in which the fire regimes are located. 

The table below shows the area of each fire regime type across the province and the proportion of each fire regime as a percentage of total area. The NA value represents those BEC units which are not assigned to a BUMO_regime, likley as they do not occur within the BUMO study area. 

```{r, echo = FALSE}

#bec_sum <- read.csv( fs::path(spatialDir, "fire_regime","bec_fr_total_ha_BC.csv"))
fr_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_BC.csv")) |> 
  select(-X) |> 
  #rename("fr_pc_bc" = fr_pc_bumo) |> 
  select(BUMO_regime, fr_area_ha, total_ha, fr_pc_bc)

knitr::kable(fr_sum)

#fr_bumo_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_bumo.csv"))

```

The table below shows the area and percentage of each fire regime type within the BUMO study area. 
<br>

```{r, echo = FALSE}

fr_bumo_sum <- read.csv(fs::path(spatialDir, "fire_regime","fr_total_ha_bumo.csv"))|> 
  select(-X) 

knitr::kable(fr_bumo_sum)

```
<br>


## 2. Fires across BC 

This plot shows the total area burnt across BC for all fire years, the number above each point indicates the total number of fires within that year. Note some early records are not accurate for fire numbers. 
<br>
<br>

```{r pressure, echo=FALSE, fig.cap="", out.width = '90%'}
p1 <- fs::path(spatialDir, "fire_regime","1_fires_across_bc.jpg")
knitr::include_graphics(p1)

```
<br>
<br>

We can review the proportion of area burnt by fire regime relative to the total area burnt in that year (figure 2), or by the proportion of area burnt by fire regime relative to the total area of a given fire regime (figure 3). I reviewed these by total area across the province, or by area within the BUMO study site. 


### By Province

The figure below shows the proportion that each fire regime contributed to all fires in the given year. 

```{r plot2, echo=FALSE, fig.cap="Figure 2", out.width = '90%'}
p2 <- fs::path(spatialDir, "fire_regime","2_fire_fr_pc_across_bc.jpg")
knitr::include_graphics(p2)

```


The figure below shows the percentage area of fire for each fire regime as a percent of the total area of the given fire regime across the province. 

```{r plot2a, echo=FALSE, fig.cap="Figure 3", out.width = '90%'}
p2a <- fs::path(spatialDir, "fire_regime","2a_fire_fr_pc__of_total_fr_across_bc.jpg")
knitr::include_graphics(p2a)

```
<br>
<br>

### By BUMO study area

We can also review the proportions of the fire regimes burn within the BUMO study area. In the figure below I show the % of area burnt by fire regime relative to total area (within BUMO) for each given fire regime type. 


```{r plot 4, echo=FALSE, fig.cap="Figure 4", out.width = '90%'}
p3 <- fs::path(spatialDir, "fire_regime","3_fire_fr_pc_across_bumo.jpg")
knitr::include_graphics(p3)
```


## 10 yr average within BuMO study area

I also estimated the 10 yr average mean centered on the year (i.e mean of 5 years prior and 4 years post). These calculations used NA to show years where data could not be estimated by this methos, however other metrics are available. Note in this figure the Y axis is scaled to each region and not static. 


```{r plot 5, echo=FALSE, fig.cap="Figure 5", out.width = '90%'}
p4 <- fs::path(spatialDir, "fire_regime","4_fire_fr_pc_across_bumo_10yr.jpg")
knitr::include_graphics(p4)
```

## 30 yr average within BuMO study area

This figure shows the 30 year rolling average for all fire regimes illustrating the average % burnt by fire regime by total area of each fire regime within BuMo study area. 


```{r plot 6, echo=FALSE, fig.cap="Figure 6", out.width = '90%'}
p5 <- fs::path(spatialDir, "fire_regime","5_fire_fr_pc_across_bumo_30yr.jpg")
knitr::include_graphics(p5)
```


All collated data (by province or BUMO study area) is output as csv for further investigations. 



<br>

<br>






## Appendices 

The key to distinguish fire regimes based on BEC zones are shown in the table below. 

```{r fire_regime key, echo=FALSE}
knitr::kable(key)

```


